Idiosyncratic, Ordinal Tax To Sustain a Digital Economy

### 2019-08-03 writing/short note on digital tax/DigitalTax.tex
adjusted paper based on Nate's feedback.

Recently, the French Senate approved a law that imposes a $3\%$ tax on revenue generated from digital services by companies above a certain size. 
While there is a lot of debate about political consequences of this action, it is actually interesting to reverse the question and instead ask what are the long-term implications of an economy with no such digital tax. 
More generally, we can see a digital service as a special case of products with low or zero cost of transportation. 
With basic economic models we show that an economy with no transportation cost is prone to monopolisation in that minuscule, random differences in quality are rewarded disproportionally. 
We then propose a distance-based tax to counter-balance the tendencies of random centralisation.  
Compared to an obvious tax that scales with physical (cardinal) distance, we show that a ranked (ordinal) distance tax leverages the benefits of digitalisation while keeping the economy stable.

@@

To analyse the potential impact of vanishing distance-based transaction costs, we consider the following model economy: 
Two firms offering the same product are competing for a total of $n$ potential buyers. 
Firms and buyers are randomly placed on a one-dimensional ring of length $\ell$ with periodic boundary conditions (Figure \ref{fig:topology}). 
Hence, the maximum distance between any buyer and firm is $D=\ell/2$.

\begin{figure}[!htb]
	\centering
	\includegraphics[width=0.3\textwidth]{topology.pdf}
	\caption{	Ring of length $\ell=1$ representing a `geography', on which we place at random two firms (circles) and $n=12$ buyers (crosses). 
			Here we have picked a balanced configuration where each firm has $n=6$ closest potential customers. 
			Throughout this article, we consider the buyer and firm arrangement fixed as shown here. 
			Any other generic arrangement or geography could be used and would not significantly alter our results.
			The two firms compete for profit.
			They set their production quantity and prices according to the Nash equilibrium, 
			taking into consideration the different levels of taxation. 
			}
	\label{fig:topology}
\end{figure}

@@

The buyers cannot buy at price $p_i$, but additionally have to pay a transaction cost $\lambda \cdot D^\gamma \cdot p_i$ where $D$ is the idiosyncratic distance between a given buyer and firm $i$. 
From hereon, we shall call this the \textit{cardinal tax}. 
In its most basic form, the gravity model predicts a value of $\gamma=2$, whereas here we have set $\gamma=1$.  
\footnote{
\label{ftn:wlog}
Changing these numbers does not alter the qualitative statements of our results, 
while some of the numbers quoted in Figure \ref{fig:results} may change.
} 
The case of zero cost of transportation is recovered in the limit where the scaling factor $\lambda$ approaches zero. 
The buyers are all characterised by the same negatively sloped demand curve, $p = u - q$ with $u$ the upper price limit at which no good is bought.
In the remainder of this article, we set $\lambda = 0.1$ and $u = 120$. \footref{ftn:wlog}

The two firms choose capacity and price to maximize their respective profit in equilibrium. 
Here, we assume they first compete on capacity, by simultaneously declaring how many units they produce;
next, they compete on price, taking into consideration the tax and assuming the location of buyers is known. 
This process, consisting of these two dependent sub-games, is similar to the game theoretical problems studied by \cite{Kreps1983}, the major difference being the idiosyncratic tax.
Quantity and price offered by each firm are determined as the Nash equilibrium of this two-stage game. 
%The firm's strategy-space is thus given by the quantity that they can produce and the prices that they can offer. 
Because the tax depends on the exact location of buyers and firms, 
it is not possible to determine the Nash equilibrium analytically, for general cases. 
Instead, we use the Gambit library developed by \cite{McKelvey2016} to find the Nash equilibrium numerically. 
%To this end, the strategy space of capacities and prices are discretised into a sufficiently fine grid. 
%The quantity-price two-step game is solved in reverse order: 
%For every choice of quantity, the equilibrium price is determined.
%Then, the equilibrium quantity is determined.

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=0.5\textwidth]{performance}
	\caption{	On the left axis (blue bars), we show the relative difference in profit of the two firms.
			If both firms have equal production cost, their revenue is similar. 
			This remains true if there is a marginal $1\%$ difference in cost, as long as there is a distance dependent cardinal tax. 
			Imbalance in revenue is increased drastically once that tax is removed, but is decreased again in case it is replaced by an ordinal tax. 
			Major imbalances in production costs (i.e. competitive advantage) lead to significantly different revenue with and without tax. 
			On the right axis (orange bars), we show the relative difference in total market revenue, as generated by the two firms combined. 
			The revenue is rescaled to a value of $1$ for the case of no tax. 
			All other revenue values are in proportion thereof. 
			While the cardinal tax significantly decreases revenue, the ordinal tax brings it back to a similar level than without any tax. 
			}
	\label{fig:results}
\end{figure}

>>>>

Recently, the French Senate approved a law that imposes a $3\%$ tax on revenue generated from digital services by companies above a certain size. 
While there is a lot of debate about political consequences of this action, it is actually interesting to reverse the question and instead ask what are the long-term implications of an economy with no such digital tax. 
More generally, we can see a digital service as a special case of products with low or zero cost of transportation. 
With basic economic models we show that an economy with no transportation costs is prone to monopolisation in that minuscule, random differences in quality are rewarded disproportionally. 
We then propose a distance-based tax to counter-balance the tendencies of random centralisation.  
Compared to an obvious tax that scales with physical (cardinal) distance, we show that a ranked (ordinal) distance tax leverages the benefits of digitalisation while keeping the economy stable.

@@

To analyse the potential impact of vanishing distance-based transaction costs, we consider the following model economy: 
Two firms offering the same product are competing for a total of $n$ potential buyers. 
Firms and buyers are randomly placed on a one-dimensional ring of length $\ell$ with periodic boundary conditions (Figure \ref{fig:topology}). 
Hence, the maximum distance between any buyer and firm is $d=\ell/2$.

\begin{figure}[!htb]
	\centering
	\includegraphics[width=0.3\textwidth]{circle_topology}
	\caption{	Ring of length $\ell=1$ representing a `geography', on which we place at random two firms (circles) and $n=12$ buyers (crosses). 
			Here we have picked a balanced configuration where each firm has $n=6$ closest potential customers. 
			Throughout this article, we consider the buyer and firm arrangement fixed as shown here. 
			Any other generic arrangement or geography could be used and would not significantly alter our results.
			The two firms compete for profit.
			They set their production quantity and prices according to the Nash equilibrium, 
			taking into consideration the different levels of taxation. 
			}
	\label{fig:topology}
\end{figure}

@@

The buyers cannot buy at price $p_i$, but additionally have to pay a transaction cost $\lambda \cdot d^\gamma \cdot p_i$ where $d$ is the distance between a given buyer and firm $i$. 
From hereon, we shall call this the \textit{cardinal tax}. 
In its most basic form, the gravity model predicts a value of $\gamma=2$, whereas here we have set $\gamma=1$.  
\footnote{
\label{ftn:wlog}
Changing these numbers does not alter the qualitative statements of our results, 
while some of the numbers quoted in Figure \ref{fig:results} may change.
} 
The case of zero cost of transportation is recovered in the limit where the scaling factor $\lambda$ approaches zero. 
The buyers are all characterised by the same negatively sloped demand curve, $p = u - q$ with $u$ the upper price limit at which no good is bought.
In the remainder of this article, we set $\lambda = 0.1$ and $u = 120$. \footref{ftn:wlog}

The two firms choose capacity and price to maximize their respective profit in equilibrium. 
Here, we assume they first compete on capacity, by simultaneously declaring how many units they produce;
next, they compete on price, taking into consideration the tax and assuming the location of buyers is known. 
This process, consisting of these two dependent sub-games, is similar to the game theoretical problems studied by \cite{Kreps1983}, the major difference being the buyer-seller specific tax.
Quantity and price offered by each firm are determined as the Nash equilibrium of this two-stage game. 
%The firm's strategy-space is thus given by the quantity that they can produce and the prices that they can offer. 
Because the tax depends on the exact location of buyers and firms, 
it is not possible to determine the Nash equilibrium analytically, for general cases. 
Instead, we use the Gambit library developed by \cite{McKelvey2016} to find the Nash equilibrium numerically. 
%To this end, the strategy space of capacities and prices are discretised into a sufficiently fine grid. 
%The quantity-price two-step game is solved in reverse order: 
%For every choice of quantity, the equilibrium price is determined.
%Then, the equilibrium quantity is determined.

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=0.5\textwidth]{performance_histogram}
	\caption{	On the left axis (blue bars), we show the relative difference in profit of the two firms.
			If both firms have equal production cost, their revenue is similar. 
			This remains true if there is a marginal $1\%$ difference in cost, as long as there is a distance dependent cardinal tax. 
			Imbalance in revenue is increased drastically once that tax is removed, but is decreased again in case it is replaced by an ordinal tax. 
			Major imbalances in production costs (i.e. competitive advantage) lead to significantly different revenue with and without tax. 
			On the right axis (orange bars), we show the relative difference in total market revenue, as generated by the two firms combined. 
			The revenue is rescaled to a value of $1$ for the case of no tax. 
			All other revenue values are in proportion thereof. 
			While the cardinal tax significantly decreases revenue, the ordinal tax brings it back to a similar level than without any tax. 
			}
	\label{fig:results}
\end{figure}
